A sample computation

### 2024-10-02 draft_systoles.tex
computed push in H_2 for SL_3(Z)

\DeclareMathOperator{\Spec}{Spec}        % Spec
\DeclareMathOperator{\Proj}{Proj}        % Proj
\DeclareMathOperator{\Reg}{Reg}        % Regulator
\DeclareMathOperator{\disc}{disc}        % discriminant

@@

\begin{Rem}
A computation shows that for $p = 100$ we have
\begin{gather*}
\alpha_1 = -9999.99999999000 , \\
\alpha_2 = -0.01000000500000625, \\
\alpha_3 = 0.00999999500000625
.\end{gather*}
It is not important that $p$ is not a prime in this case as the estimate works for any sufficiently large $p$.
\end{Rem}

@@

\section{From $\SL$ to $\SLn$}

@@

\bibliography{mybib}
\bibliographystyle{plain}

>>>>

\DeclareMathOperator{\Spec}{Spec}        % Spec
\DeclareMathOperator{\Proj}{Proj}        % Proj
\DeclareMathOperator{\Reg}{Reg}        % Regulator
\DeclareMathOperator{\disc}{disc}        % discriminant
\DeclareMathOperator{\Hom}{Hom}        % Hom
\DeclareMathOperator{\Aut}{Aut}        % Aut

@@

\begin{Rem}
	\label{rem_roots_estimate}
A computation shows that for $p = 100$ we have
\begin{gather*}
\alpha_1 = -9999.99999999000 , \\
\alpha_2 = -0.01000000500000625, \\
\alpha_3 = 0.00999999500000625
.\end{gather*}
It is not important that $p$ is not a prime in this case as the estimate works for any sufficiently large $p$.
\end{Rem}

@@

\section{Computing (co)homology}
Christophe Soulé \cite{soule1978}
computed the cohomology of $\SL$
to be
\begin{table}[h!]
	\centering
\begin{tabular}{|c|c|c|c|c|c|c}
	\hline
	$n$        & $0$                               & $1$                               & $2$                               & $3$                               & $4$  & \ldots\\
\hline 
	$H^n(\SL, \mathbb{Z})$ & $\mathbb{Z}$ & $0$ & $0$ & $(\mathbb{Z}/2\mathbb{Z})^{2} $ & $(\mathbb{Z}/3\mathbb{Z})^{2} \oplus (\mathbb{Z}/4\mathbb{Z})^{2} $ & \ldots \\
\hline
\end{tabular}.
\end{table}

The universal coefficient formula gives us the homology.
\begin{table}[h!]
	\centering
\begin{tabular}{|c|c|c|c|c|c}
	\hline
	$n$        & $0$                               & $1$                               & $2$                               & $3$                               & \ldots\\
\hline 
	$H_n(\SL, \mathbb{Z})$ & $\mathbb{Z}$ & $0$ &  $(\mathbb{Z}/2\mathbb{Z})^{2} $ & $(\mathbb{Z}/3\mathbb{Z})^{2} \oplus (\mathbb{Z}/4\mathbb{Z})^{2} $ & \ldots \\
\hline
\end{tabular}
\end{table}

Recall the matrices 
\[
A = 
\begin{pmatrix}
0 & 0 & 1 \\
1 & 0 & 0 \\
0 & 1 & -p^2
\end{pmatrix}
,
B = \Id + p A =
\begin{pmatrix}
1 & 0 & p \\
p & 1 & 0 \\
0 & p & 1 - p^3
\end{pmatrix}
.\] 
We denote $\tilde A$ by $B$ for convenience.

As $A, B $ commute they define a map
\[
f: \mathbb{Z}^2 \to \SL, 
\] 
\[
(k, l) \mapsto A^k B^l
.\]

\begin{Thm}
	\label{thm_push_homology_vanishes}
The pushforward map in the second homology 
\[
f_*:
H_2(\mathbb{Z}^2, \mathbb{Z}) \cong \mathbb{Z} \to 
H_2(\SL, \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2} 
\] 
is zero.
\end{Thm}
The goal of this section is to prove this theorem.

The first observation is that one can pass from the homology to the cohomology with coefficients $\mathbb{Z}/ 2 \mathbb{Z}$.

\begin{Lem}
	\label{lemma_push_vanishes_iff_pull_vanishes}
The pushforward
\[
f_*:
H_2(\mathbb{Z}^2, \mathbb{Z}) \cong \mathbb{Z} \to 
H_2(\SL, \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2} 
\] 
is zero 
iff
the pullback
\[
f^*:
H^2(\SL,  \mathbb{Z}/2 \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2} \to
H^2(\mathbb{Z}^2, \mathbb{Z}/2 \mathbb{Z}) \cong \mathbb{Z}/2 \mathbb{Z}
\] 
is zero.
\end{Lem}
\begin{proof}
As the first homology 
\[
H_1(\mathbb{Z}^2, \mathbb{Z}) \cong \mathbb{Z}^2, 
H_1(\SL, \mathbb{Z}) \cong 0
\] 
are torsion-free, the universal coefficien formula gives perfect $\mathbb{Z}/2\mathbb{Z}$-pairings between homology and cohomology
\[
H^2(\mathbb{Z}^2, \mathbb{Z}/2 \mathbb{Z}) = \Hom(H_2(\mathbb{Z}^2, \mathbb{Z}), \mathbb{Z}/2 \mathbb{Z})
,\] 
\[
H^2(\SL, \mathbb{Z}/2 \mathbb{Z}) = \Hom(H_2(\SL, \mathbb{Z}), \mathbb{Z}/2 \mathbb{Z})
.\] 
Therefore the pullback in cohomology is the transposed map of the pushforward in homology and one is zero iff the other is zero.
\end{proof}

Next, let us find a nicer description for $H^2(\SL, \mathbb{Z}/2 \mathbb{Z})$.

\begin{Cla}
\begin{itemize}
	\item The second cohomology $H^2(G, M)$ classify the extensions $0 \to M \to \tilde G \to G \to 0$;
	\item In particular, $H^2(G, \mathbb{Z}/2\mathbb{Z})$ correspond to the extensions of degree $2$;
\item Given an extension $0 \to M \to \tilde G \to G \to 0$ and a principle $G$-bundle over $X$ we get its class in $H^2(X, M)$ via the boundary (in other words, Bockstein) homomorphism $H^1(X, G) \to H^2(X, M)$.  In details, the Bockstein goes by taking the $1$-cocycle defining the $G-$bundle to $\tilde G$ and apply the Cech differential;
	\item $H^2( \mathbf{SL}_3(\mathbb{R}), \mathbb{Z}/ 2 \mathbb{Z}) \cong
	H^2( \mathbf{SL}_3(\mathbb{F}_2), \mathbb{Z}/ 2 \mathbb{Z}) \cong \mathbb{Z}/2\mathbb{Z}$;
\item The corresponding unique extensions of degree two are $\widetilde{\mathbf{SL}_3}(\mathbb{R})$ and $\mathbf{SL}_2(\mathbb{F}_7)$.
	We make use of the exceptional isomorphism $\mathbf{PSL}_2(\mathbb{F}_7) \cong \mathbf{SL}_3(\mathbb{F}_2)$, see, for example, \cite{merzon_psl_2_7_sl_3_2}.
\item If $E$ is a principle $\mathbf{SL}_3(\mathbb{R})$-bundle on $X$ the corresponding class in $H^2(X, \mathbb{Z}/ 2 \mathbb{Z})$ is called $w_2(E)$, the second Stiefel-Whitney class;
\item If $E$ is a principle $\mathbf{SL}_3(\mathbb{F}_2)$-bundle on $X$ let us denote the corresponding class in $H^2(X, \mathbb{Z}/ 2 \mathbb{Z})$ by $u_2(E)$.
\end{itemize}
\end{Cla}
This is all very well known.
For the cohomology of $\mathbf{SL}_3(\mathbb{F}_2)$ see, for example, \cite{adem2006lecturescohomologyfinitegroups}.

Now, consider the natural embeddings
\[
i_1: \SL \embeds {\mathbf{SL}_3(\mathbb{R})}
, \] 
\[
i_2: \SL \embeds {\mathbf{SL}_3(\mathbb{F}_2)}
.\]

\begin{Prop}
The pullback 
\[
i_1^* \oplus i_2^*: 
H^2(B\mathbf{SL}_3(\mathbb{R}), \mathbb{Z}/ 2 \mathbb{Z}) \oplus
H^2(B\mathbf{SL}_3(\mathbb{F}_2), \mathbb{Z}/ 2 \mathbb{Z}) \cong (\mathbb{Z} / 2 \mathbb{Z})^2 \to 
H^2(\SL, \mathbb{Z}/ 2 \mathbb{Z}) \cong (\mathbb{Z}/2\mathbb{Z})^{2}
\] 
is an isomorphism.
\end{Prop}

\begin{proof}
It is enough to show that a principal $\SL$ bundle $E$ on $\mathbb{T}^2$ can have an arbitrary combination of $(w_2(E), u_2(E)) \in (\mathbb{Z}/ 2\mathbb{Z})^2$.

Plan: we shall construct principal bundles $E_1, E_2$ on $\mathbb{T}^2$ with $(w_2(E_1), u_2(E_1)) = (1, 0)$ and $u_2(E_2) = 1$.
This will prove the proposition.

The first bundle $E_1$ is defined via the monodromy matrices $C_1, C_2$
\[
C_1 = 
\begin{pmatrix}
-1 & 0 & 0 \\
0 & -1 & 0 \\
0 & 0 & 1
\end{pmatrix}
,
C_2 =
\begin{pmatrix}
1 & 0 & 0 \\
0 & -1 & 0 \\
0 & 0 & -1
\end{pmatrix} \in \SL
.\] 
By the splitting principle we have
\[
w(E_1)
= (1 + x) (1 + x + y) (1 + y) 
= 1 + xy
,\] 
where $x, y$ denote the generators of $H^1(\mathbb{T}^2, \mathbb{Z}/ 2 \mathbb{Z})$.
Therefore $w_2(E_1) = 1$.
As modulo $2$ the matrices $C_1, C_2$ are the identity the bundle $E_1$ mod $2$ is trivial; thus $u_2(E_1) = 0$.

Define $D_1' = 
\left( \begin{smallmatrix}0&6\\1&0\end{smallmatrix}\right) 
, D_2' =
\left( \begin{smallmatrix}2&3\\3&5\end{smallmatrix}\right)
\in \mathbf{SL}_2(\mathbb{F}_7)$.
Project the matrices to 
$\in \mathbf{PSL}_2(\mathbb{F}_7) \cong \in \mathbf{SL}_3(\mathbb{F}_2) $.
Pick a pair of lifts $D_1, D_2 \in \mathbf{SL}_3(\mathbb{Z})$.

Define the bundle $E_2$ corresponding to the monodromy matrices $D_1, D_2$. 
We claim that $u_2(E_2) = 1$.

Indeed, we shall compute the Bockstein morphism $H^1(\mathbb{T}^2, \mathbf{SL}_3(\mathbb{F}_2) \to H^2(\mathbb{T}^2, \mathbb{Z}/ 2 \mathbb{Z})$ by lifting the cocycles.

One checks that $D_1' D_2' = - D_2' D_1'$; thus the image of the Bockstein map is nontrivial.

Thus we have showed that $E_1, E_2$ have the right characteristic classes and we are done.
\end{proof}
\begin{proof}[Proof of the Theorem \ref{thm_push_homology_vanishes}]
By Lemma \ref{lemma_push_vanishes_iff_pull_vanishes} we shall show that the pullback map in second $\mathbb{Z}/ 2 \mathbb{Z}$-cohomology vanishes.

Consider the principal $\SL$-bundle $E$ on $\mathbb{T}^2$ associated with $A, B$.
We should show that $w_2(E) = 0$ and $u_2(E) = 0$.

By the splitting principle
using that the roots of the characteristic polynomial 
\ref{rem_roots_estimate} 
satisfy \[
\alpha_1, 1 + p \alpha_1 < 0, 
\alpha_2, 1 + p \alpha_2 < 0, 
\alpha_3, 1 + p \alpha_3 > 0 
\] 
we get the Stiefel-Whitney class to be
\[
w(E) = w(E_1) w(E_2) w (E_3)
= (1 + x + y) (1 + x + y) (1) 
= 1
,\]
and thus $w_2(E) = 0$ indeed does vanish.

Modulo $2$ one easily spots that $B = A^5 \pmod{2}$, thus their lifts to $\mathbf{SL}_2(\mathbf{F}_7)$ do commute and the image of the Bockstain morphism vanishes; thus $u_2(E) = 0$.
% the explanation is that we should consider SL_3(F_2) in the automorphisms of F_8 = F_2[t]/t^3 + t^2 + 1 as a linear space over F_2; then A is multiplication by t and B is a multiplication by 1 + t = t^5. 
\end{proof}

\newpage
\section{From $\SL$ to $\SLn$}

@@

\bibliography{mybib}
\bibliographystyle{alpha}
